# Panel Data - One Area

## Aim

We are given a dataset containing daily counts of diseases from one geographical area. We want to identify:

- Does seasonality exist?
- If seasonality exists, when are the high/low seasons?
- Is there a general yearly trend (i.e. increasing or decreasing from year to year?)
- Is daily rainfall associated with the number of cases?

```{r warning=FALSE}
library(data.table)
library(ggplot2)
set.seed(4)

AMPLITUDE <- 1.5
SEASONAL_HORIZONTAL_SHIFT <- 20

d <- data.table(date=seq.Date(
  from=as.Date("2000-01-01"),
  to=as.Date("2018-12-31"),
  by=1))
d[,date:=as.Date(date,origin="1970-01-1")]
d[,year:=as.numeric(format.Date(date,"%G"))]
d[,week:=as.numeric(format.Date(date,"%V"))]
d[,month:=as.numeric(format.Date(date,"%m"))]
d[,yearMinus2000:=year-2000]
d[,dailyrainfall:=runif(.N, min=0, max=10)]

d[,dayOfYear:=as.numeric(format.Date(date,"%j"))]
d[,seasonalEffect:=sin(2*pi*(dayOfYear-SEASONAL_HORIZONTAL_SHIFT)/365)]
d[,mu := exp(0.1 + yearMinus2000*0.1 + seasonalEffect*AMPLITUDE)]
d[,y:=rpois(.N,mu)]
```

## Data

Here we show the true data, and note that there is an increasing annual trend (the data gets higher as time goes on) and there is a seasonal pattern (one peak/trough per year)
```{r}
q <- ggplot(d,aes(x=date,y=y))
q <- q + geom_line(lwd=0.25)
q <- q + scale_x_date("Time")
q <- q + scale_y_continuous("Cases")
q
```

We split out the data for a few years and see a clear seasonal trend:

```{r}
q <- ggplot(d[year %in% c(2005:2010)],aes(x=dayOfYear,y=y))
q <- q + facet_wrap(~year)
q <- q + geom_point()
q <- q + stat_smooth(colour="red")
q <- q + scale_y_continuous("Day of year")
q <- q + scale_y_continuous("Cases")
q
```

## Model With Non-Parametric Seasonality

If we want to investigate the seasonality of our data, and identify when are the peaks and troughs, we can use non-parametric approaches. They are flexible and easy to implement, but they can lack power and be hard to interpret:

- Create a categorical variable for the seasons (e.g. `spring`, `summer`, `autumn`, `winter`) and include this in the regression model
- Create a categorical variable for the months (e.g. `Jan`, `Feb`, ..., `Dec`) and include this in the regression model

```{r}
nfit0 <- glm(y~yearMinus2000 + dailyrainfall, data=d, family=poisson())
nfit1 <- glm(y~yearMinus2000 + dailyrainfall + as.factor(month), data=d, family=poisson())
```

We can test the `month` categorical variable using a likelihood ratio test:

```{r}
lmtest::lrtest(nfit0, nfit1)
```

And then we can look at the output of our regression:

```{r}
summary(nfit1)
```

*NOTE:* See that this is basically the same as a normal regression.

If we want to identify outbreaks, then we need to use the standard prediction interval formula:

$$
95\% \text{ CI} = \text{sample average} \pm 1.96 \times \text{sample standard deviation} \sqrt{ 1 + 1 / n}
$$
This allows us to identify what the expected thresholds are:

```{r}
pred <- predict(nfit1, type = "response", se.fit = T, newdata = d)
d[, threshold0 := pred$fit]
d[, threshold2 := FarringtonThreshold(pred, phi = 1, z = 2, skewness.transform = "2/3")]
```

```{r}
q <- ggplot(d[year>2015],aes(x=date,y=y))
q <- q + geom_ribbon(mapping=aes(ymin=-Inf,ymax=threshold2),fill="green",alpha=0.5)
q <- q + geom_ribbon(mapping=aes(ymin=threshold2,ymax=Inf),fill="red",alpha=0.5)
q <- q + geom_line(lwd=0.25)
q <- q + geom_point(data=d[year>2015 & y>threshold2],colour="black",size=2.5)
q <- q + geom_point(data=d[year>2015 & y>threshold2],colour="red",size=1.5)
q <- q + scale_x_date("Time")
q <- q + scale_y_continuous("Cases")
q
```

## Model With Parametric Seasonality 

Parametric approaches are more powerful but require more effort:

- Identify the periodicity of the seasonality (how many days between peaks?)
- Using trigonometry, transform `day of year` into variables that appropriately model the observed periodicity
- Obtain coefficient estimates
- Back-transform these estimates into human-understandable values (day of peak, day of trough)

*NOTE:* You don't always have to investigate seasonality! It depends entirely on what the purpose of your analysis is!

The Lomb-Scargle Periodogram shows a clear seasonality with a period of 365 days.

```{r}
# R CODE 
lomb::lsp(d$y,from=100,to=500,ofac=1,type="period")
```

We then generate two new variables `cos365` and `sin365` and perform a likelihood ratio test to see if they are significant or not. This is done with two simple poisson regressions.

```{r}
# R CODE
d[,cos365:=cos(dayOfYear*2*pi/365)]
d[,sin365:=sin(dayOfYear*2*pi/365)]

pfit0 <- glm(y~yearMinus2000 + dailyrainfall, data=d, family=poisson())
pfit1 <- glm(y~yearMinus2000 + dailyrainfall + sin365 + cos365, data=d, family=poisson())
```

We can test the seasonality using a likelihood ratio test (which we already strongly suspected due to the periodogram):

```{r}
lmtest::lrtest(pfit0, pfit1)
```

And then we can look at the output of our regression:

```{r}
summary(pfit1)
```

We also see that the (significant!) coefficient for `year` is `0.1` which means that for each additional year, the outcome increases by `exp(0.1)=1.11`. We also see that the coefficient for `dailyrainfall` was not significant, which means that we did not find a significant association between the outcome and `dailyrainfall`.

*NOTE:* See that this is basically the same as a normal regression.

Through the likelihood ratio test we saw a clear significant seasonal effect. We can now use trigonometry to back-calculate the amplitude and location of peak/troughs from the `cos365` and `sin365` estimates:
```{r}
b1 <- 1.428417 # sin coefficient
b2 <- -0.512912 # cos coefficient
amplitude <- sqrt(b1^2 + b2^2)
p <- atan(b1/b2) * 365/2/pi
if (p > 0) {
    peak <- p
    trough <- p + 365/2
} else {
    peak <- p + 365/2
    trough <- p + 365
}
if (b1 < 0) {
    g <- peak
    peak <- trough
    trough <- g
}
print(sprintf("amplitude is estimated as %s, peak is estimated as %s, trough is estimated as %s",round(amplitude,2),round(peak),round(trough)))

print(sprintf("true values are: amplitude: %s, peak: %s, trough: %s",round(AMPLITUDE,2),round(365/4+SEASONAL_HORIZONTAL_SHIFT),round(3*365/4+SEASONAL_HORIZONTAL_SHIFT)))

```

*NOTE:* An amplitude of 1.5 means that when comparing the average time of year to the peak, the peak is expected to be `exp(1.5)=4.5` times higher than average. We take the exponential because we have run a poisson regression (so think incident rate ratio).

## Autocorrelation

We check the `pacf` of the residuals to ensure that there is no autocorrelation. If we observe autocorrelation in our residuals, then we need to use a `robust` variance estimator (i.e. it makes our estimated variances bigger to account for our poor model fitting).

Here we see that our non-parametric seasonality model has not accounted for all of the associations in the data, so there is some autocorrelation in the residuals:

```{r}
d[,residuals:=residuals(nfit1, type = "response")]
d[,predicted:=predict(nfit1, type = "response")]
pacf(d$residuals)
```

Here we see that our parametric seasonality model has  accounted for all of the associations in the data, so there is no autocorrelation in the residuals:

```{r}
d[,residuals:=residuals(pfit1, type = "response")]
d[,predicted:=predict(pfit1, type = "response")]
pacf(d$residuals)
```
