---
output: html_document
editor_options: 
  chunk_output_type: console
---

# Panel data - Multiple Areas

## Aim

We are given a dataset containing yearly counts of diseases from multiple geographical areas.

We will explore how this is different from the one area case. 

For this section, we will use linear regression to make the calculations easier to work with, but the principle is the same as with poisson and negative binomial regression.

## Creating the data

```{r warning=FALSE}
library(data.table)
library(lme4)
library(ggplot2)
set.seed(4)

fylkeIntercepts <- data.table(fylke = 1:3, fylkeIntercepts = c(30, 0, -30))

d <- data.table(fylke = rep(1:3, each = 100))
d <- merge(d, fylkeIntercepts, by = "fylke")
d[, mainIntercept := 5]
d[, x := runif(.N)]
d[, year := sample(c(1950:2020), .N, replace = T)]
d[, yearMinus1950 := year - 1950]
d[, mu := mainIntercept + fylkeIntercepts + 0.5 * yearMinus1950]
d[, y := round(rnorm(.N, mu, sd = 1))]
# d[,y:=rpois(.N,mu)]

d[fylke == 1 & !year %in% c(1950:1980), y := NA]
d[fylke == 2 & !year %in% c(1980:2000), y := NA]
d[fylke == 3 & !year %in% c(2000:2020), y := NA]

d <- na.omit(d)
```

\newpage

## Investigating the data

We begin by blindly looking at the data.

```{r}
q <- ggplot(d, aes(x = year, y = y))
q <- q + geom_point()
q <- q + scale_x_continuous("Year")
q <- q + scale_y_continuous("Cases")
q
```

We then blindly run a linear regression and we see that the number of cases is decreasing over time.

```{r}
q <- ggplot(d, aes(x = year, y = y))
q <- q + geom_point()
q <- q + stat_smooth(method = "lm", se = F, colour = "red")
q <- q + scale_x_continuous("Year")
q <- q + scale_y_continuous("Cases")
q
```

However, once we reveal the geographical clustering in the data, we can see that our regression model is extremely wrong: 

```{r}
q <- ggplot(d, aes(x = year, y = y))
q <- q + geom_point(mapping = aes(colour = as.factor(fylke)))
q <- q + stat_smooth(method = "lm", se = F, colour = "red")
q <- q + scale_x_continuous("Year")
q <- q + scale_y_continuous("Cases")
q <- q + scale_colour_brewer("Area", palette = "Set1")
q
```

**Note:** The differing `number of cases` in each geographical area could also be due to differing `populations`. Always remember to `use a denominator`!!!

## Time Trend Within Geographical Areas

What we actually want to do, is observe the time trends `within` each geographical area:

```{r}
fit <- lm(y ~ as.factor(fylke) + yearMinus1950, data = d)
d[, p := predict(fit, newdata = d)]

q <- ggplot(d, aes(x = year, y = y))
q <- q + geom_point(mapping = aes(colour = as.factor(fylke)))
q <- q + geom_line(mapping = aes(y = p, group = as.factor(fylke)), colour = "red")
q <- q + scale_x_continuous("Year")
q <- q + scale_y_continuous("Cases")
q <- q + scale_colour_brewer("Area", palette = "Set1")
q
```

So how do we achieve this? Basically, by identifying how the areas differ from each other, and then raising/lowering each area as appropriate. The amount that each area differs from the global average is called the `area-specific intercept`.

Once we have removed the `area-specific intercept` from each geographical area, we can (basically) just run a normal regression to find the overall time trend.

```{r}
d[, yMinusIntercept := y]
d[fylke == 2, yMinusIntercept := y - -30.0944]
d[fylke == 3, yMinusIntercept := y - -59.6795]

q <- ggplot(d, aes(x = year, y = yMinusIntercept))
q <- q + geom_point(mapping = aes(colour = as.factor(fylke)))
q <- q + stat_smooth(method = "lm", se = F, colour = "red")
q <- q + scale_x_continuous("Year")
q <- q + scale_y_continuous("Cases Minus Area-Specific Intercept")
q <- q + scale_colour_brewer("Area", palette = "Set1")
q
```

## Applying The Regression Models

So how do you obtain `area-specific intercepts`? If you have a large amount of datapoints for each area, and not many areas, you can just include `area` as a categorical variable in your regression. These `area-specific intercepts` are called `fixed effects`.

```{r}
fit <- lm(y ~ as.factor(fylke) + yearMinus1950, data = d)
summary(fit)
```

If you have few datapoints for each area, and a lot of areas, you will need to use a `mixed effects` regression model, including `random intercepts` for each `area`. 

```{r}
fit <- lme4::lmer(y ~ yearMinus1950 + (1 | fylke), data = d)
summary(fit)

lme4::ranef(fit)
```

## Hints For Future Analyses

In general:

$$\text{number of observations} = \text{number of time points} \times \text{number of areas}$$

So 5 years of data for aggregated Norway is worse than 5 years of data for every municipality in Norway.



