--- 
title: "Time Series and Longitudinal Analysis"
author: "Richard White"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: folkehelseinstituttet/surveillance_timeseries
description: "A short course in learning which statistical methods should be applied where."
---

# Preface {-}

When dealing with data measured over time, there are two kinds of analyses that can be performed.

"Time series" analyses generally deal with one variable (the outcome). We can then either:

1. Predict the future only using the previous observations. E.g. predict tomorrow's temperature, using today's and yesterday's temperature as exposures. We will not be focusing on these kinds of analyses.
2. Estimate descriptive statistics about the data. E.g. Today's data is much higher than expected (outbreak?). We will focus on these kinds of analyses.

If we have more than one variable measured over time (e.g. outcome and an exposure) then we can run regression analyses. E.g. seeing how the number of tuberculosis patients (outcome) is affected by the number of immigrants to Norway (exposure) over a 20 year period. We will focus on these kinds of analyses.

It is important to note that if we define our exposure as "time" then we can use the regression framework to estimate descriptive statistics about the data. This means we can use the same regression framework for the two kinds of analyses we will be focusing on.

The "regression framework" is very similar to ordinary regressions that you have been working with for many years. The only difference is that some of the data **may** have more advanced data structures that your normal methods cannot handle.

<!--chapter:end:index.Rmd-->

# Definitions and Scenarios

## Panel Data

Panel data is a set of data with measurements repeated at equally spaced points. For example, number of influenza cases recorded every day, or every week, or every year would be considered panel data. The number of influenza cases on Jan 31, Feb 3, and Nov 21 in 2018 would not be considered panel data.

## Autocorrelation

When you have panel data, autocorrelation is the correlation between subsequent observations. For example, if you have daily observations, then the 1 day autocorrelation is the correlation between observations 1 day apart, and likewise the 2 day autocorrelation is the correlation between observations 2 days apart.

## Scenarios

In this course we will consider two scenarios where we have multiple observations for each geographical area:

- Panel data: One geographical area, with/without autocorrelation
- Multiple geographical areas without autocorrelation

Note, the following scenario can be covered by standard regression models:

- Multiple geographical areas, one time point/observation per geographical area

## Useful Code

This code is used to calculate prediction intervals. In its most basic form it is:

$$
95\% \text{ prediction interval} = \text{sample average} \pm 1.96 \times \text{sample standard deviation} \sqrt{ 1 + 1 / n}
$$
However, due to the skewness of the count data, we often choose to use a `2/3s transformation`.

```{r}
sykdomspuls::FarringtonThreshold
```

Please note that a prediction interval is not the same as a confidence interval!

<!--chapter:end:01-reference.Rmd-->

# Panel Data - One Area

## Aim

We are given a dataset containing daily counts of diseases from one geographical area. We want to identify:

1. Does seasonality exist?
2. If seasonality exists, when are the high/low seasons?
3. Is there a general yearly trend (i.e. increasing or decreasing from year to year?)
4. Is daily rainfall associated with the number of cases?
5. When are there outbreaks?

```{r warning=FALSE}
library(data.table)
library(ggplot2)
set.seed(4)

AMPLITUDE <- 1.5
SEASONAL_HORIZONTAL_SHIFT <- 20

d <- data.table(date=seq.Date(
  from=as.Date("2000-01-01"),
  to=as.Date("2018-12-31"),
  by=1))
d[,date:=as.Date(date,origin="1970-01-1")]
d[,year:=as.numeric(format.Date(date,"%G"))]
d[,week:=as.numeric(format.Date(date,"%V"))]
d[,month:=as.numeric(format.Date(date,"%m"))]
d[,yearMinus2000:=year-2000]
d[,dailyrainfall:=runif(.N, min=0, max=10)]

d[,dayOfYear:=as.numeric(format.Date(date,"%j"))]
d[,seasonalEffect:=sin(2*pi*(dayOfYear-SEASONAL_HORIZONTAL_SHIFT)/365)]
d[,mu := exp(0.1 + yearMinus2000*0.1 + seasonalEffect*AMPLITUDE)]
d[,y:=rpois(.N,mu)]
```

## Data

Here we show the true data, and note that there is an increasing annual trend (the data gets higher as time goes on) and there is a seasonal pattern (one peak/trough per year)
```{r}
q <- ggplot(d,aes(x=date,y=y))
q <- q + geom_line(lwd=0.25)
q <- q + scale_x_date("Time")
q <- q + scale_y_continuous("Cases")
q
```

We split out the data for a few years and see a clear seasonal trend:

```{r message=FALSE, warning=FALSE}
q <- ggplot(d[year %in% c(2005:2010)],aes(x=dayOfYear,y=y))
q <- q + facet_wrap(~year)
q <- q + geom_point()
q <- q + stat_smooth(colour="red")
q <- q + scale_x_continuous("Day of year")
q <- q + scale_y_continuous("Cases")
q
```

## Model With Non-Parametric Seasonality

If we want to investigate the seasonality of our data, and identify when are the peaks and troughs, we can use non-parametric approaches. They are flexible and easy to implement, but they can lack power and be hard to interpret:

- Create a categorical variable for the seasons (e.g. `spring`, `summer`, `autumn`, `winter`) and include this in the regression model
- Create a categorical variable for the months (e.g. `Jan`, `Feb`, ..., `Dec`) and include this in the regression model

```{r}
nfit0 <- glm(y~yearMinus2000 + dailyrainfall, data=d, family=poisson())
nfit1 <- glm(y~yearMinus2000 + dailyrainfall + as.factor(month), data=d, family=poisson())
```

### Seasonality

We can test the `month` categorical variable using a likelihood ratio test:

```{r}
lmtest::lrtest(nfit0, nfit1)
```

**Question 1:** Does seasonality exist?

And then we can look at the output of our regression:

```{r}
summary(nfit1)
```

*NOTE:* See that this is basically the same as a normal regression.

**Question 2:** If seasonality exists, when are the high/low seasons?

### Yearly trend

**Question 3:** Is there a general yearly trend (i.e. increasing or decreasing from year to year?)

### Association With Rainfall

**Question 4:** Is daily rainfall associated with the number of cases?

### Outbreaks

If we want to identify outbreaks, then we need to use the standard prediction interval formula:

$$
95\% \text{ prediction interval} = \text{sample average} \pm 1.96 \times \text{sample standard deviation} \sqrt{ 1 + 1 / n}
$$
This allows us to identify what the expected thresholds are:

```{r}
pred <- predict(nfit1, type = "response", se.fit = T, newdata = d)
d[, threshold0 := pred$fit]
d[, threshold2 := sykdomspuls::FarringtonThreshold(pred, phi = 1, z = 2, skewness.transform = "2/3")]
```

**Question 5:** When are there outbreaks?

```{r}
q <- ggplot(d[year>2015],aes(x=date,y=y))
q <- q + geom_ribbon(mapping=aes(ymin=-Inf,ymax=threshold2),fill="green",alpha=0.5)
q <- q + geom_ribbon(mapping=aes(ymin=threshold2,ymax=Inf),fill="red",alpha=0.5)
q <- q + geom_line(lwd=0.25)
q <- q + geom_point(data=d[year>2015 & y>threshold2],colour="black",size=2.5)
q <- q + geom_point(data=d[year>2015 & y>threshold2],colour="red",size=1.5)
q <- q + scale_x_date("Time")
q <- q + scale_y_continuous("Cases")
q
```

## Model With Parametric Seasonality 

Parametric approaches are more powerful but require more effort:

- Identify the periodicity of the seasonality (how many days between peaks?)
- Using trigonometry, transform `day of year` into variables that appropriately model the observed periodicity
- Obtain coefficient estimates
- Back-transform these estimates into human-understandable values (day of peak, day of trough)

*NOTE:* You don't always have to investigate seasonality! It depends entirely on what the purpose of your analysis is!

### Seasonality

The Lomb-Scargle Periodogram shows a clear seasonality with a period of 365 days.

```{r}
# R CODE 
lomb::lsp(d$y,from=100,to=500,ofac=1,type="period")
```

We then generate two new variables `cos365` and `sin365` and perform a likelihood ratio test to see if they are significant or not. This is done with two simple poisson regressions.

```{r}
# R CODE
d[,cos365:=cos(dayOfYear*2*pi/365)]
d[,sin365:=sin(dayOfYear*2*pi/365)]

pfit0 <- glm(y~yearMinus2000 + dailyrainfall, data=d, family=poisson())
pfit1 <- glm(y~yearMinus2000 + dailyrainfall + sin365 + cos365, data=d, family=poisson())
```

We can test the seasonality using a likelihood ratio test (which we already strongly suspected due to the periodogram):

```{r}
lmtest::lrtest(pfit0, pfit1)
```

**Question 1:** Does seasonality exist?

And then we can look at the output of our regression:

```{r}
summary(pfit1)
```

We also see that the (significant!) coefficient for `year` is `0.1` which means that for each additional year, the outcome increases by `exp(0.1)=1.11`. We also see that the coefficient for `dailyrainfall` was not significant, which means that we did not find a significant association between the outcome and `dailyrainfall`.

*NOTE:* See that this is basically the same as a normal regression.

Through the likelihood ratio test we saw a clear significant seasonal effect. We can now use trigonometry to back-calculate the amplitude and location of peak/troughs from the `cos365` and `sin365` estimates:

```{r}
RAWmisc::TransformCosSinToAmplitudePeakTrough
```

```{r}
retval <- RAWmisc::TransformCosSinToAmplitudePeakTrough(
  cos_b = -0.512912, # cos coefficient
  sin_b = 1.428417 # sin coefficient
  )

print(sprintf("amplitude is estimated as %s",round(retval$amplitude,2)))
print(sprintf("peak is estimated as %s",round(retval$peak)))
print(sprintf("trough is estimated as %s",round(retval$trough)))

print(sprintf("true amplitude is %s",round(AMPLITUDE,2)))
print(sprintf("true peak is %s",round(365/4+SEASONAL_HORIZONTAL_SHIFT)))
print(sprintf("true trough is %s",round(3*365/4+SEASONAL_HORIZONTAL_SHIFT)))
```

*NOTE:* An amplitude of 1.5 means that when comparing the average time of year to the peak, the peak is expected to be `exp(1.5)=4.5` times higher than average. We take the exponential because we have run a poisson regression (so think incident rate ratio).

**Question 2:** If seasonality exists, when are the high/low seasons?

### Yearly trend

**Question 3:** Is there a general yearly trend (i.e. increasing or decreasing from year to year?)

### Association With Rainfall

**Question 4:** Is daily rainfall associated with the number of cases?

## Autocorrelation

We check the `pacf` of the residuals to ensure that there is no autocorrelation. If we observe autocorrelation in our residuals, then we need to use a `robust` variance estimator (i.e. it makes our estimated variances bigger to account for our poor model fitting).

Here we see that our non-parametric seasonality model has not accounted for all of the associations in the data, so there is some autocorrelation in the residuals:

```{r}
d[,residuals:=residuals(nfit1, type = "response")]
d[,predicted:=predict(nfit1, type = "response")]
pacf(d$residuals)
```

Here we see that our parametric seasonality model has  accounted for all of the associations in the data, so there is no autocorrelation in the residuals:

```{r}
d[,residuals:=residuals(pfit1, type = "response")]
d[,predicted:=predict(pfit1, type = "response")]
pacf(d$residuals)
```

## Hints For Future Analyses

### Zeroes In Your Individial -> Aggregated Dataset

If your data is at the individual level (i.e. one row per case), then you will need to aggregate it to daily/weekly/monthly levels. If your data source is a registry, and you assume that your dataset contains all of the reported cases, then this means `"no reports"="no cases"`. This means that after aggregating, **you need to make sure that your collapsed/aggregated dataset has zeroes in it!!**

In the following dataset, we have one case per day from `2000-01-01` until `2000-01-29` and then again one case per day from `2000-08-08` until `2000-12-31`.

```{r}
d <- data.table(date=seq.Date(
  from=as.Date("2000-01-01"),
  to=as.Date("2000-12-31"),
  by=1))
d <- d[-c(30:220)]
d[,id:=1:.N]
d[,month:=as.numeric(format.Date(date,"%m"))]

print(d)
```

If we collapse the data into months:

```{r}
collapsed <- d[,.(
  n=.N
), keyby=.(month)]

print(collapsed)
```

You see that we are missing months `1` through to `6`. We cannot analyse this data, because **we do not have any zeros**.

How do we fix this? We create a `skeleton` of our results:

```{r}
skeleton <- data.table(month=1:12)
print(skeleton)
```

We then merge our collapsed data with the skeleton:

```{r}
final <- merge(skeleton, collapsed, by="month", all.x=TRUE)
print(final)
```

We then set all of the "missing" to 0:

```{r}
final[is.na(n), n:=0]
print(final)
```

Now we can analyze our data!

### Lagging Of Exposures

If you are interested in seeing how exposures affect your outcome (e.g. `rainfall`) then you might want to consider lagging your exposure. This will show you `how did the rainfall from last week affect the number of cases this week?`






<!--chapter:end:02-one-area.Rmd-->

---
output: html_document
editor_options: 
  chunk_output_type: console
---

# Not panel data: Multiple areas

## Aim

We are given a dataset containing counts of diseases from multiple geographical areas. We want to identify:

- Is there a general yearly trend (i.e. increasing or decreasing from year to year?)
- Is variable `x` associated with the outcome?

\newpage

## Creating the data

The data for this chapter is available at: http://rwhite.no/longitudinal_analysis/data/chapter_5.csv

```{r warning=FALSE}
library(data.table)
library(lme4)
set.seed(4)

fylkeIntercepts <- data.table(fylke=1:20,fylkeIntercepts=rnorm(20))

d <- data.table(fylke=rep(1:20,each=100))
d <- merge(d,fylkeIntercepts,by="fylke")
d[,mainIntercept:=3]
d[,x:=runif(.N)]
d[,year:=sample(c(1950:2018),.N,replace=T)]
d[,mu := exp(mainIntercept + fylkeIntercepts + 3*x)]
d[,y:=rpois(.N,mu)]

fwrite(d,"data/chapter_5.csv")
```

\newpage

## Investigating the data

We can see from the data that we have 20 geographical areas (`fylke`) with 100 observations for each fylke, but the sampling did not happen consistently (some years have multiple measurements, other years have no measurements).

This means we have:

- multiple geographical areas
- multiple observations in each geographical area
- not panel data

```{r}
print(d)
```

\newpage

## Regression

For this scenario, we use the `lme4::glmer` function in R. We need to introduce a `(1|fylke)` term to identify the geographical areas (i.e. clusters). In STATA we use the `meglm` function and introduce a `|| fylke:` term to identify the geographical areas (i.e. clusters).

```{}
// STATA CODE STARTS
insheet using "chapter_5.csv", clear

gen yearMinus2000 = year-2000
meglm y x yearMinus2000 || fylke:, family(poisson)
// STATA CODE ENDS
```

```{r warning=FALSE}
# R CODE
d[,yearMinus2000:=year-2000]
summary(fit <- lme4::glmer(y~x + yearMinus2000 + (1|fylke),data=d,family=poisson()))
```

You can see that the format of the results is the same as an ordinary regression.



<!--chapter:end:03-multiple-areas-no-panel-data.Rmd-->

